One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-12 paper/results.tex
update
intro draft completed
minor updates

\section{Results}
We first present results from FARGO simulations. The 2D disc spans
$[R_\mathrm{min}, R_\mathrm{max}] = [0.4,10]R_0$. This gives a total
disc mass $M_{d}=0.086M_*$. The mass within
$R\in[R_\mathrm{min},R_{1}]$ is $0.017M_*$, that within
$R\in[R_{1},R_{2}]$ is $0.049M_*$, and that within
$R\in[R_{2},R_\mathrm{max}]$ is $0.021M_*$. We use a resolution of
$N_R\times N_\phi = 1024\times 2048$, or about $16$ grids per $H$, and
adopt $\epsilon_g=10^{-4}hR$ for the   
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}.

@@

\subsection{Reference run with random perturbations}
To obtain a picture of the overall disc evolution, we describe a 
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
Fig. \ref{fargo_modeamp} plots evolution of the maximum 
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to saturation of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_DZ_fargo}
  \caption{Evolution of maxima in non-axisymmetric surface density 
    in the fiducial FARGO simulation initialised with perturbations
    with $m\in[1,10]$.\label{fargo_modeamp}} 
\end{figure}

\begin{figure}
  \includegraphics[scale=0.27]{figures/polarxy_dens050}\includegraphics[scale=0.27,clip=true,trim=2.26cm 
  0cm 0cm 
  0cm]{figures/polarxy_dens110}\includegraphics[scale=0.27,clip=true,trim=2.26cm
  0cm 0cm 0cm]{figures/polarxy_dens130} 
  \caption{Visualisation of the fiducial FARGO 2D simulation in
    Fig. \ref{fargo_modeamp}. The total  
    non-axisymmetric surface density
    $\Delta\Sigma$ is shown. \label{fargo_2d}} 
\end{figure}

Fig. \ref{2d_angmom} shows the evolution of disc angular momentum
components. Only the $m=0,\,1$ components are 
plotted since they are dominant. The $m=1$ structure has
an associated negative angular momentum.  
Its growth is compensated by an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 + \Delta
J_1 \sim 0$. Note that FARGO does not conserve angular momentum
exactly. However, we find the total angular momentum varies by 
$|\Delta J/J|= O(10^{-6})$, and is much smaller in magnitude than the
change in the angular momenta components, $|\Delta J_{0,1}/J|>
O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
is transferred from the one-arm spiral to the background disc.

\begin{figure}
  % scale=0.41
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
  \caption{Evolution of angular momentum components in the fiducial
    FARGO simulation in Fig. \ref{fargo_modeamp}---\ref{fargo_2d}. The
    perturbation relative to $t=0$ in 2D is shown in units of the 
    initial total angular momentum $J_\mathrm{ref}$.\label{2d_angmom}} 
\end{figure}

\subsection{Trapping the low-frequency $m=1$ spiral}\label{fargo_m1}
In order to focus on the $m=1$ spiral that eventually emerges, from
now on we examine simulations initialised with
$M=1$ in Eq. \ref{randpert}. The disc 
evolution is largely similar to the previous case, but the final spiral is
more coherent, which facilitates the analysis below.

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-arm spiral
we measure a co-rotation radius and growth rate of 
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p\lesssim 0.3\Omega$ in $R\in[R_{1},R_{2}]$,
where it is  most prominent. The growth rate $\gamma$ is also slow
relative to the local rotation. (Although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$.) Thus, the spiral 
pattern is nearly stationary.

@@

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We measure $k$ numerically and plot
its normalised value in Fig. \ref{fargo_wavenumber}. We find 
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
can apply results from local theory self-consistently. Recall $k>0$
corresponds to a trailing spiral, as seen in Fig. \ref{2d_fargo_viz}.

@@

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. %Note that, with the above measured wavenumber $k$, 
%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
%sign (long waves) to be taken. 
Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
region where the $m=1$ spiral develops. This suggests that the one-arm
spiral is trapped. Note in this region, $\nu^2 - 1 +
Q^{-2}\simeq 0.1\ll 1$, as required for consistency with
Eq. \ref{dispersion} (with the lower sign taken, i.e. long waves).  
There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the  
spiral disturbance in the inner disc \citep{lin11b}. 
%this would be a de-stabilising
% effect.  we discuss this in discussions?

@@

\subsection{Dependence on the imposed temperature profile} 
At this point we hypothesize the growth of the $m=1$ spiral mode is
associated with the imposed temperature gradient. The torque
responsible for its growth, Eq. \ref{baroclinic_torque}, is
proportional to $q$, the power-law index used to set the sound-speed
profile (Eq. \ref{sound-speed}). To test this, we carried out
simulations with $q\in[0,1]$ using only an $m=1$ initial 
perturbation. However, to keep the Toomre profiles similar, we adjust
the surface density power-law index such that $s = (3+q)/2$ for each
case presented below. %should not be important, local effects

>>>>

\section{Results}\label{results2d}
We first present results from FARGO simulations. The 2D disc spans
$[R_\mathrm{min}, R_\mathrm{max}] = [0.4,10]R_0$. This gives a total
disc mass $M_{d}=0.086M_*$. The mass within
$R\in[R_\mathrm{min},R_{1}]$ is $0.017M_*$, that within
$R\in[R_{1},R_{2}]$ is $0.049M_*$, and that within
$R\in[R_{2},R_\mathrm{max}]$ is $0.021M_*$. We use a resolution of
$N_R\times N_\phi = 1024\times 2048$, or about $16$ grids per $H$, and
adopt $\epsilon_g=10^{-4}hR$ for the   
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}.

@@

\subsection{Reference run with random perturbations}
To obtain a picture of the overall disc evolution, we describe a 
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
Fig. \ref{fargo_modeamp} plots evolution of the maximum 
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to non-linear interaction of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_DZ_fargo}
  \caption{Evolution of non-axisymmetric surface density maxima 
    in the FARGO simulation initialised with perturbations
    with $m\in[1,10]$.\label{fargo_modeamp}} 
\end{figure}

\begin{figure}
  \includegraphics[scale=0.27]{figures/polarxy_dens050}\includegraphics[scale=0.27,clip=true,trim=2.26cm 
  0cm 0cm 
  0cm]{figures/polarxy_dens110}\includegraphics[scale=0.27,clip=true,trim=2.26cm
  0cm 0cm 0cm]{figures/polarxy_dens130} 
  \caption{Visualisation of the FARGO 2D simulation in
    Fig. \ref{fargo_modeamp}. The total  
    non-axisymmetric surface density
    $\Delta\Sigma$ is shown. \label{fargo_2d}} 
\end{figure}

Fig. \ref{2d_angmom} shows the evolution of disc angular momentum
components. Only the $m=0,\,1$ components are 
plotted since they are dominant. The $m=1$ structure has
an associated negative angular momentum.  
Its growth is compensated by an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 + \Delta
J_1 \sim 0$. Note that FARGO does not conserve angular momentum
exactly. However, we find the total angular momentum varies by 
$|\Delta J/J|= O(10^{-6})$, and is much smaller than the
change in the angular momenta components, $|\Delta J_{0,1}/J|>
O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
is transferred from the one-arm spiral to the background disc.

\begin{figure}
  % scale=0.41
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
  \caption{Evolution of angular momentum components in the 
    FARGO simulation in Fig. \ref{fargo_modeamp}---\ref{fargo_2d}. The
    perturbation relative to $t=0$ in 2D is shown in units of the 
    initial total angular momentum $J_\mathrm{ref}$.\label{2d_angmom}} 
\end{figure}

\subsection{Trapping the low-frequency $m=1$ spiral}\label{fargo_m1}
In order to focus on the $m=1$ spiral that eventually emerges, from
now on we examine simulations initialised 
$M=1$ in Eq. \ref{randpert}. The disc 
evolution is largely similar to the previous case, but the one-armed spiral is
more coherent, which facilitates the analysis below.

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-armed spiral
we measure a co-rotation radius and growth rate  
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p\lesssim 0.3\Omega$ in $R\in[R_{1},R_{2}]$,
where it is  most prominent. The growth rate $\gamma$ is also slow
relative to the local rotation. (Although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$.) Thus, the spiral 
pattern is nearly stationary.

@@

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We measured $k$ numerically and plot
its normalised value in Fig. \ref{fargo_wavenumber}. We find 
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
can apply results from local theory
self-consistently. Fig. \ref{2d_fargo_viz} shows the $m=1$ spiral is
trailing, which is consistent with $k>0$.

@@

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. %Note that, with the above measured wavenumber $k$, 
%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
%sign (long waves) to be taken. 
Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
region where the $m=1$ spiral develops. This suggests that the one-arm
spiral is trapped. Note in this region, $\nu^2 - 1 +
Q^{-2}\simeq 0.1\ll 1$, as required for consistency with
Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, acoustic waves may be launched in $R\gtrsim 7.2R_0$ by
the spiral disturbance in the inner disc \citep{lin11b}. 
%implications: effect on dynamics in outer disc

@@

\subsection{Dependence on the imposed temperature profile} 
At this point we hypothesize the growth of the $m=1$ spiral mode is
associated with the imposed temperature gradient. The torque
responsible for its growth, Eq. \ref{baroclinic_torque}, is
proportional to $q$, the power-law index used to set the sound-speed
profile (Eq. \ref{sound-speed}). To test this, we carried out
simulations with $q\in[0,1]$ using only an $m=1$ initial 
perturbation. However, to keep the Toomre profiles similar, we adjust
the surface density power-law index such that $s = (3+q)/2$.
